### 2023-02-28 let.tex

% TODO standardize naming: marking vs mark insertion, marks vs error marks vs markings
% TODO uniform style for referencing rules

\section{The Marking Calculus}
\label{sec:calculus}

Programmers often work with program states that are syntactically well-formed but not well-typed.
\Cref{fig:calculus-examples} shows common type errors as they appear in the Hazel programming
environment. For example, a simple error that may occur is the use of an unbound variable. As in
\cref{fig:calculus-examples-unbound}, editor services indicate the usage of unbound $y$ is invalid
by highlighting it in red.

\input{figs/calculus-examples}

Sometimes, the type an expression \emph{should} be is known. Consider
\cref{fig:calculus-examples-inconsistent-types}, in which it is expected that both operands of the
$+$ operator are numbers. Since $\textsf{true}$ is not a number, a type error arises, and Hazel
indicates a mismatch between the expected and actual types of the operand. The application in
\cref{fig:calculus-examples-app-non-lambda} of a non-lambda leads to a similar kind of error. In
this case, however, the expectation is not any singular type, but any member of a family of function
types. \Cref{fig:calculus-examples-inconsistent-branches} presents an error of inconsistent
branches: the branches of the if-else expression are of different types. Hazel opts to highlight the
entire expression in such a situation, while other tools may choose one branch to be ``correct'' and
indicate a type mismatch error on the other.

% TODO elaborate on this part

Unfortunately, formal typing semantics usually only give meaning to well-typed programs, leaving
such ill-typed ones meaningless. This is insufficient for editor services and language servers,
which must provide users feedback on where and why a program is ill-typed. Moreover, they must
continue providing editor services even when an error is encountered. Indeed, similarly to
error-recovering parsers, real-world type checkers must recover from type errors. Traditionally,
this has been performed in an \emph{ad hoc} manner [citations?], with different languages and tools
adopting varying approaches in the localization, reportage, and recovery from errors.

% TODO examples of this ad hoc stuff

The key contribution of this section is a formal calculus, called the \emph{marking calculus}, for
type error localization in bidirectional systems, a judgmental framework that provides a unified
theory for the design of typing semantics that supports ill-typed programs.
\Cref{sec:calculus-calculus} describes its form and metatheory by example on a minimal gradually
typed lambda calculus. All rules and theorems may be found in the supplementary appendix, alongside
a complete mechanization in the Agda proof assistant, which is discussed in
\cref{sec:calculus-agda}. Finally, in \cref{sec:calculus-poly}, we consider how the system might be
extended to more complex typing features, such as System F-style polymorphism.

% TODO not really sure when to start talking about bidirectional typing

\emph{Bidirectional typing} combines type checking, which determines if a program satisfies a given
type, and type synthesis, which generates a type from the program. This gives a simple algorithmic
system in which type information is recursively propagated through terms and allows for local type
inference.

% TODO talk about idea that bidirectional typing makes error localization easy? i.e. why choose
% bidirectional?

\subsection{The Calculus}
\label{sec:calculus-calculus}

% TODO outline the structure of this section at the beginning
% TODO make it clear unmarked/marked languages share same types

We now formally introduce the \emph{marking calculus} through its application to a gradually typed
lambda calculus extended with numbers and booleans. Consider the syntax for such a language, given
as $\EMName$ in \cref{fig:calculus-syntax}. The base types $\TNum$ and $\TBool$ classify number and
boolean expressions. The number literal corresponding to the mathematical number $n$ is given by
$\ENumMV$, and there is a single addition operation on numeric expressions. $\ETrue$ and $\EFalse$
correspond to the boolean values $\textsf{true}$ and $\textsf{false}$, respectively, and
$\EIf{\EMV_1}{\EMV_2}{\EMV_3}$ gives a conditional structure. In addition to these forms,
$\TUnknown$ is the unknown type from gradual type theory.

% TODO include holes
\begin{figure}[htbp]
  \[\begin{array}{rrcl}
    \TMName  & \TMV  & \Coloneqq & \TUnknown \mid \TNum \mid \TBool \mid \TArrow{\TMV}{\TMV} \\
    \EMName  & \EMV  & \Coloneqq & x \mid \ELam{x}{\TMV}{\EMV} \mid \EAp{\EMV}{\EMV}
                       \mid           \ENumMV \mid \EPlus{\EMV}{\EMV}
                       \mid           \ETrue \mid \EFalse \mid \EIf{\EMV}{\EMV}{\EMV} \\
    \ECMName & \ECMV & \Coloneqq & x \mid \ECLam{x}{\TMV}{\ECMV} \mid \ECAp{\ECMV}{\ECMV}
                       \mid           \ECNumMV \mid \ECPlus{\ECMV}{\ECMV}
                       \mid           \ECTrue \mid \ECFalse \mid \ECIf{\ECMV}{\ECMV}{\ECMV} \\
             &       & \mid         & \ECUnbound{x} \mid \ECInconType{\ECMV} \mid \ECInconBr{\ECMV}{\ECMV}{\ECMV}
  \end{array}\]
  \caption{Syntax of the extended gradually typed lambda calculus}
  \label{fig:calculus-syntax}
\end{figure}

Types classify expressions by a \emph{bidirectional type system}, which employs two mutually defined
judgments. \emph{Type synthesis}, written $\ctxSynType{\ctx}{\EMV}{\TMV}$, establishes that, under
the typing context $\ctx$, the expression $e$ synthesises or infers the type $\TMV$. \emph{Type
analysis}, written $\ctxAnaType{\ctx}{\EMV}{\TMV}$, states that the expression $\EMV$ may appear
where an expression of type $\TMV$ is expected. We may give these typing semantics traditionally,
without considering error localization and recovery.

The \emph{marking calculus} builds atop these typing semantics into a two-layer system:
%
\begin{itemize}
  \item The \textbf{unmarked language}, which is the original language. Expressions of this language
    are called \emph{unmarked expressions}, denoted by the metavariable $\EMV$.

  \item The \textbf{marked language}, which mirrors the structure of the unmarked language but is
    extended with explicit \emph{error marks}. We call expressions of this language to be
    \emph{marked expressions}, denoted $\ECMV$.
    % TODO more intuition for what marked expressions are?

  \item \textbf{Marking}, which \emph{marks}, i.e. type checks and localizes errors for, any unmarked
    expression into a marked expression, inserting error marks where appropriate.
\end{itemize}
%
In other words, we extend the ordinary typing semantics of a language with a secondary ``marked''
language and ``mark'' unmarked programs into marked ones. This corresponds to a type checking
process with error localization and recovery. The marked language may then serve as a foundation for
other semantic services. % such as THI.

% TODO more intuition

\Cref{fig:calculus-syntax} furthermore gives the syntax for marked expressions (as $\ECMName$),
which replicates all the forms of unmarked expressions. There are additionally three kinds of
\emph{error marks}, which represent the different type errors that may arise.
%
\begin{itemize}
  \item $\ECUnbound{x}$, the \emph{unbound variable mark}, indicates the use of an unbound variable
    $x$, as in \cref{fig:calculus-examples-unbound}.

  \item $\ECInconType{\ECMV}$, the \emph{inconsistent type mark}, indicates that the type of an
    expression does not match an expected type.

  \item $\ECInconBr{\ECMV_1}{\ECMV_2}{\ECMV_3}$, the \emph{inconsistent branches mark}, indicates
    that the branches of a conditional have inconsistent types.
\end{itemize}
%
Throughout the remainder of this section, as we consider the formulation of marking, we will
motivate and give precise semantics for these error marks. Note that this is a non-exhaustive list;
depending on the syntactic forms and typing features found in the language, other kinds of marks may
be necessary. Indeed, in \cref{sec:calculus-poly}, we consider the kinds of marks that may be
required in an language with explicit polymorphism.

The marked language possesses its own typing semantics, also formulated bidirectionally. We write
$\ctxSynType{\ctx}{\ECMV}{\TMV}$ for synthesis and $\ctxAnaType{\ctx}{\ECMV}{\TMV}$ for analysis.
Finally, marking is bidirectional as well. The synthetic marking judgment
$\ctxSynFixedInto{\ctx}{\EMV}{\ECMV}{\TMV}$ establishes that, under the context $\Gamma$, the
unmarked expression $\EMV$ is ``marked into'' the marked expression $\ECMV$, which synthesizes type
$\TMV$. Analogously, the analytic marking judgment $\ctxAnaFixedInto{\ctx}{\EMV}{\ECMV}{\TMV}$
states that $\EMV$ is marked into $\ECMV$, which analyses against $\TMV$.

How can we ensure that the marking procedure is correctly defined? There are two critical
metatheorems to keep in mind as we continue. The first is a \emph{totality} of marking:
%
\begin{theorem}[name=Marking Totality] \
  \label{thm:calculus-marking-totality}
  \begin{enumerate}
    \item For all $\ctx$ and $\EMV$, there exist $\ECMV$ and $\TMV$ such that
      $\ctxSynFixedInto{\ctx}{\EMV}{\ECMV}{\TMV}$.
  \item For all $\ctx$, $\EMV$, and $\TMV$, there exists $\ECMV$ such that
    $\ctxAnaFixedInto{\ctx}{\EMV}{\ECMV}{\TMV}$.
  \end{enumerate}
\end{theorem}
%
That is, we may mark \emph{any} syntactically well-formed program.

Furthermore, marking should preserve the syntactic structure of programs, modulo error marks. Thus,
we define a notion of \emph{mark erasure}, whose definition is given
\cref{fig:calculus-mark-erasure}, which converts marked expressions back into unmarked expression by
removing error marks. Then, \cref{thm:calculus-marking-well-formedness} provides a
\emph{well-formedness} criterion marking.
%
\begin{theorem}[name=Marking Well-Formedness] \
  \label{thm:calculus-marking-well-formedness}
  \begin{enumerate}
    \item If $\ctxSynFixedInto{\ctx}{\EMV}{\ECMV}{\TMV}$, then $\ctxSynType{\ctx}{\ECMV}{\TMV}$ and
      $\erasesTo{\ECMV}{\EMV}$.
    \item If $\ctxAnaFixedInto{\ctx}{\EMV}{\ECMV}{\TMV}$, then $\ctxAnaType{\ctx}{\ECMV}{\TMV}$ and
      $\erasesTo{\ECMV}{\EMV}$.
  \end{enumerate}
  % TODO split this into two theorems?
  % TODO not really sure how to intuitively describe the part about types matching up
\end{theorem}
%
In other words, if an unmarked expression is marked into some marked expression, erasing error marks
from the marked expression gives the original unmarked expression. We also ensure that marking gives
well-typed marked expressions; indeed, their types should match the types synthesised or analysed
against in marking. % TODO more intuition for this theorem

\input{figs/calculus-mark-erasure}

In the remainder of this subsection, we consider the different type errors that may arise in the
unmarked language; in this way, the typing semantics of the unmarked language guide the construction
of the marking process.

\subsubsection{Numbers}
\label{sec:calculus-numbers}

% TODO need to say something about subsumption and how it will be handled later.
To start, let us consider the simple case of numbers. Because of subsumption, we need only define a
synthesis rule for unmarked numbers, in which numbers synthesize the type $\TNum$:
%
\begin{mathpar}
  \judgment{ }{
    \ctxSynType{\ctx}{\ENumMV}{\TNum}
  }{USNum}
\end{mathpar}

Into what should numbers be marked? Quite straightforwardly, we simply give the same number as a
marked expression, synthesizing again $\TNum$. No type errors may occur in a single number, so we
only need the following rules for marking unmarked numbers and typing marked numbers:
%
\begin{mathpar}
  \judgment{
  }{
    \ctxSynFixedInto{\ctx}{\ENumMV}{\ECNumMV}{\TNum}
  }{ISNum}

  \judgment{ }{
    \ctxSynType{\ctx}{\ECNumMV}{\TNum}
  }{MSNum}
\end{mathpar}

In addition expressions, the type of both operands should be $\TNum$. In a bidirectional system, we
thus analyse the operands against $\TNum$, giving the following rule:
%
\begin{mathpar}
  \judgment{
    \ctxAnaType{\ctx}{\EMV_1}{\TNum} \\
    \ctxAnaType{\ctx}{\EMV_2}{\TNum}
  }{
    \ctxSynType{\ctx}{\EPlus{\EMV_1}{\EMV_2}}{\TNum}
  }{USPlus}
\end{mathpar}

The marking rule mirrors \textsc{USPlus} closely. Since a known, expected type for each operand is
known, we shift the ``blame'' for any type errors to them. Hence, we recursively mark each operand
in analytic mode and rebuild the marked addition expression. The marked typing rule then mirrors
\textsc{USPlus} exactly.
%
\begin{mathpar}
  \judgment{
    \ctxAnaFixedInto{\ctx}{\EMV_1}{\ECMV_1}{\TNum} \\\\
    \ctxAnaFixedInto{\ctx}{\EMV_2}{\ECMV_2}{\TNum}
  }{
    \ctxSynFixedInto{\ctx}{\EPlus{\EMV_1}{\EMV_2}}{\ECPlus{\ECMV_1}{\ECMV_2}}{\TNum}
  }{ISPlus}

  \judgment{
    \ctxAnaType{\ctx}{\ECMV_1}{\TNum} \\\\
    \ctxAnaType{\ctx}{\ECMV_2}{\TNum}
  }{
    \ctxSynType{\ctx}{\ECPlus{\ECMV_1}{\ECMV_2}}{\TNum}
  }{MSPlus}
\end{mathpar}

\subsubsection{Variables}
\label{sec:calculus-variables}

Let us now consider the more interesting case of variables. The typing of unmarked variables in the
gradually typed lambda calculus is standard. Again, because of subsumption, there is only a
synthetic rule:
\[%
  \judgment{
    \inCtx{\ctx}{x}{\TMV}
  }{
    \ctxSynType{\ctx}{x}{\TMV}
  }{USVar}
\]%
If $x$ is bound to a type in the typing context, it synthesizes that type. Straightforwardly,
marking converts the unmarked $x$ into the same marked $x$ via the following marking and typing
rules:
%
\begin{mathpar}
  \judgment{
    \inCtx{\ctx}{x}{\TMV}
  }{
    \ctxSynFixedInto{\ctx}{x}{x}{\TMV}
  }{ISVar}

  \judgment{
    \inCtx{\ctx}{x}{\TMV}
  }{
    \ctxSynType{\ctx}{x}{\TMV}
  }{MSVar}
\end{mathpar}
% TODO more intuition on why this makes sense?

But what happens a variable, such as $y$ in \cref{fig:calculus-examples-unbound}, is \emph{not}
bound, i.e. that the premise fails? Then, \textsc{USVar} does not apply, and the typing semantics
say nothing about such expressions. When devising marking, to satisfy totality, such a possibility
motivates an \emph{unbound variable mark} $\ECUnbound{x}$, accompanied by the following rules:
%
\begin{mathpar}
  \judgment{
    \notInCtx{\ctx}{x}
  }{
    \ctxSynFixedInto{\ctx}{x}{\ECUnbound{x}}{\TUnknown}
  }{ISUnbound}

  \judgment{
    \notInCtx{\ctx}{x}
  }{
    \ctxSynType{\ctx}{\ECUnbound{x}}{\TUnknown}
  }{MSUnbound}
\end{mathpar}
%
\textsc{ISUnbound} states that an variable unbound in the context is marked into that variable
inside an unbound variable mark. \textsc{MSUnbound} then gives the typing of unbound variable marks.
Since nothing may be said about their types, we synthesize the unknown type, which allows type
checking to continue---instead of simply failing---and the unbound variable to be used in any
expression. % TODO clarify why this is correct

\subsubsection{Lambda Abstractions}
\label{sec:calculus-lambda-abstractions}

There are explicit synthesis and analysis rules for unmarked lambda abstractions:
%
\begin{mathpar}
  \judgment{
    \ctxSynType{\extendCtx{\ctx}{x}{\TMV}}{\EMV}{\TMV_2}
  }{
    \ctxSynType{\ctx}{\ELam{x}{\TMV_1}{\EMV}}{\TArrow{\TMV_1}{\TMV_2}}
  }{USLam}

  \judgment{
    \matchedArrow{\TMV_3}{\TMV_1}{\TMV_2} \\
    \consistent{\TMV}{\TMV_1} \\
    \ctxAnaType{\extendCtx{\ctx}{x}{\TMV}}{\EMV}{\TMV_2}
  }{
    \ctxAnaType{\ctx}{\ECLam{x}{\TMV}{\EMV}}{\TMV_3}
  }{UALam}
\end{mathpar}
%
The synthesis rule is standard. In the premises of \textsc{UALam}, we use the notion of
\emph{type consistency} from gradual type theory, which defines a reflexive and symmetric (but not
transitive) relation between types, writing $\consistent{\TMV_1}{\TMV_2}$ to mean that $\TMV_1$ is
consistent with $\TMV_2$. This replaces the notion of \emph{type equality} to relate the unknown
type to all other types. The analysis rule also employs the judgment
$\matchedArrow{\TMV}{\TMV_1}{\TMV_2}$, which establishes that $\TMV$ is a \emph{matched arrow type},
i.e. it may be considered an arrow type. This notion is purely a technical mechanism to avoid
duplication of rules related to arrow types. \Cref{fig:calculus-consistency-matched-arrow} gives the
rules for both judgments.

% TODO mention explicitly that by inductive hypothesis, the marking of body cannot fail?
% TODO more intuition on these two judgments
% TODO more intuition on USLam and UALam

\begin{figure}[htbp]
  \raggedright
  \judgbox{\ensuremath{\consistent{\TMV_1}{\TMV_2}}} $\TMV_1$ is consistent with $\TMV_2$
  %
  \begin{mathpar}
    \judgment{ }{
      \consistent{\TUnknown}{\TMV}
    }{TCUnknown1}

    \judgment{ }{
      \consistent{\TMV}{\TUnknown}
    }{TCUnknown2}

    \judgment{ }{
      \consistent{\TMV}{\TMV}
    }{TCRefl}

    \judgment{
      \consistent{\TMV_1}{\TMV_1'} \\
      \consistent{\TMV_2}{\TMV_2'} \\
    }{
      \consistent{\TArrow{\TMV_1}{\TMV_2}}{\TArrow{\TMV_1'}{\TMV_2'}}
    }{TCArr} \\
  \end{mathpar}

  \judgbox{\ensuremath{\matchedArrow{\TMV}{\TMV_1}{\TMV_2}}} $\TMV$ has matched arrow type $\TArrow{\TMV_1}{\TMV_2}$
  %
  \begin{mathpar}
    \judgment{ }{
      \matchedArrow{\TUnknown}{\TUnknown}{\TUnknown}
    }{TMAHole}

    \judgment{ }{
      \matchedArrow{\TArrow{\TMV}{\TMV}}{\TMV}{\TMV}
    }{TMAArr}
  \end{mathpar}
  %
  \caption{Type consistency and matched arrow types.}
  \label{fig:calculus-consistency-matched-arrow}
\end{figure}

The synthesis rule for marking intuitively follows \textsc{USLam} closely. We should recursively
mark the body with an extended context binding $x$ to $\TMV$ and construct a new marked lambda
abstraction:
%
\begin{mathpar}
  \judgment{
    \ctxSynFixedInto{\extendCtx{\ctx}{x}{\TMV_1}}{\EMV}{\ECMV}{\TMV_2}
  }{
    \ctxSynFixedInto{\ctx}{\ELam{x}{\TMV_1}{\EMV}}{\ELam{x}{\TMV_1}{\ECMV}}{\TArrow{\TMV_1}{\TMV_2}}
  }{ISLam}

  \judgment{
    \ctxSynType{\extendCtx{\ctx}{x}{\TMV}}{\ECMV}{\TMV_2}
  }{
    \ctxSynType{\ctx}{\ECLam{x}{\TMV_1}{\ECMV}}{\TArrow{\TMV_1}{\TMV_2}}
  }{MSLam}
\end{mathpar}
%
Similarly, we construct corresponding analytic rules:
%
\begin{mathpar}
  \judgment{
    \matchedArrow{\TMV_3}{\TMV_1}{\TMV_2} \\
    \consistent{\TMV}{\TMV_1} \\\\
    \ctxAnaFixedInto{\extendCtx{\ctx}{x}{\TMV}}{\EMV}{\ECMV}{\TMV_2}
  }{
    \ctxAnaFixedInto{\ctx}{\ELam{x}{\TMV}{\EMV}}{\ECLam{x}{\TMV}{\ECMV}}{\TMV_3}
  }{IALam1}

  \judgment{
    \matchedArrow{\TMV_3}{\TMV_1}{\TMV_2} \\
    \consistent{\TMV}{\TMV_1} \\\\
    \ctxAnaType{\extendCtx{\ctx}{x}{\TMV}}{\ECMV}{\TMV_2}
  }{
    \ctxAnaType{\ctx}{\ECLam{x}{\TMV}{\ECMV}}{\TMV_3}
  }{MALam1}
\end{mathpar}
%
However, recalling totality, we look again to the premises of \textsc{UALam} to see what type errors
may arise. First, what if $\TMV_3$ is not a matched arrow type? This would be the case if it were
$\TNum$, for example, and the lambda abstraction would be inconsistent with the expected type. Thus,
to the syntax we add an \emph{inconsistent type mark} $\ECInconType{\ECMV}$, which indicates that
the expression does not synthesize a type consistent with the expected type. We then have the
following rules, in which the lambda abstraction is placed inside the inconsistent type mark.
%
\begin{mathpar}
  \judgment{
    \notMatchedArrow{\TMV_3} \\
    \ctxAnaFixedInto{\extendCtx{\ctx}{x}{\TMV}}{\EMV}{\ECMV}{\TUnknown}
  }{
    \ctxAnaFixedInto{\ctx}{\ELam{x}{\TMV}{\EMV}}{\ECInconType{\ECLam{x}{\TMV}{\ECMV}}}{\TMV_3}
  }{IALam2}

  \judgment{
    \notMatchedArrow{\TMV_3} \\
    \ctxAnaType{\extendCtx{\ctx}{x}{\TMV}}{\ECMV}{\TUnknown}
  }{
    \ctxAnaType{\ctx}{\ECInconType{\ECLam{x}{\TMV}{\ECMV}}}{\TMV_3}
  }{MALam2}
\end{mathpar}
%
Though no expected output type is known, we still need to check and mark the body; it is thus
analysed against the unknown type in \textsc{IALam2}.

A similar situation arises when $\matchedArrow{\TMV_3}{\TMV_1}{\TMV_2}$, but
$\inconsistent{\TMV}{\TMV_1}$, i.e. the expected input type is inconsistent with the actual input
type, giving the last pair of analytic rules for lambda abstractions:
%
\begin{mathpar}
  \judgment{
    \matchedArrow{\TMV_3}{\TMV_1}{\TMV_2} \\
    \inconsistent{\TMV}{\TMV_1} \\\\
    \ctxAnaFixedInto{\extendCtx{\ctx}{x}{\TMV}}{\EMV}{\ECMV}{\TMV_2}
  }{
    \ctxAnaFixedInto{\ctx}{\ELam{x}{\TMV}{\EMV}}{\ECInconType{\ECLam{x}{\TMV}{\ECMV}}}{\TMV_3}
  }{IALam3}

  \judgment{
    \matchedArrow{\TMV_3}{\TMV_1}{\TMV_2} \\
    \inconsistent{\TMV}{\TMV_1} \\\\
    \ctxAnaType{\extendCtx{\ctx}{x}{\TMV}}{\ECMV}{\TMV_2}
  }{
    \ctxAnaType{\ctx}{\ECInconType{\ECLam{x}{\TMV}{\ECMV}}}{\TMV_3}
  }{MALam3}
\end{mathpar}

\subsubsection{Applications}
\label{sec:calculus-applications}

The case of application is somewhat interesting. In the unmarked language, only a synthesis rule is
necessary:
\[%
  \judgment{
    \ctxSynType{\ctx}{\EMV_1}{\TMV} \\
    \matchedArrow{\TMV}{\TMV_1}{\TMV_2} \\
    \ctxAnaType{\ctx}{\EMV_2}{\TMV_1}
  }{
    \ctxSynType{\ctx}{\EAp{\EMV_1}{\EMV_2}}{\TMV_2}
  }{USAp}
\]%
Following the same methodology, we can straightforwardly give the following marking and typing
rules:
%
\begin{mathpar}
  \judgment{
    \ctxSynFixedInto{\ctx}{\EMV_1}{\ECMV_1}{\TMV} \\
    \matchedArrow{\TMV}{\TMV_1}{\TMV_2} \\\\
    \ctxAnaFixedInto{\ctx}{\EMV_2}{\ECMV_2}{\TMV_1} \\
  }{
    \ctxSynFixedInto{\ctx}{\EAp{\EMV_1}{\EMV_2}}{\ECAp{\ECMV_1}{\ECMV_2}}{\TMV_2}
  }{ISAp1}

  \judgment{
    \ctxSynType{\ctx}{\ECMV_1}{\TMV} \\
    \matchedArrow{\TMV}{\TMV_1}{\TMV_2} \\\\
    \ctxAnaType{\ctx}{\ECMV_2}{\TMV_1}
  }{
    \ctxSynType{\ctx}{\ECAp{\ECMV_1}{\ECMV_2}}{\TMV_2}
  }{MSAp1}
\end{mathpar}

Again, to satisfy totality, we must consider the case when $\TMV$ is not a matched arrow type, such
as in the example of \cref{fig:calculus-examples-app-non-lambda}. There is no expected type for the
argument, so we perform analytic marking on $\EMV_2$ against the unknown type. The applied
expression is placed inside the inconsistent type mark, which is in accordance with the meaning of
that mark.
%
\begin{mathpar}
  \judgment{
    \ctxSynFixedInto{\ctx}{\EMV_1}{\ECMV_1}{\TMV} \\
    \notMatchedArrow{\TMV} \\\\
    \ctxAnaFixedInto{\ctx}{\EMV_2}{\ECMV_2}{\TUnknown}
  }{
    \ctxSynFixedInto{\ctx}{\EAp{\EMV_1}{\EMV_2}}{\ECApNonMatched{\ECMV_1}{\ECMV_2}}{\TUnknown}
  }{ISAp2}

  \judgment{
    \ctxSynType{\ctx}{\ECMV_1}{\TMV} \\
    \notMatchedArrow{\TMV} \\\\
    \ctxAnaType{\ctx}{\ECMV_2}{\TUnknown}
  }{
    \ctxSynType{\ctx}{\ECApNonMatched{\ECMV_1}{\ECMV_2}}{\TUnknown}
  }{MSAp2}
\end{mathpar}

It is worth noting, however, that $\EMV_1$ is not actually in analytic position, but in a synthetic
position where any arrow type is expected. Treating such ``constrained synthetic'' cases specially,
we may instead designate a specialized mark, written $\ECApNonMatchedAltAlt{\ECMV}$. This yields the
following alternative rules:
%
\[\begin{array}{rrcl}
  \ECMName & \ECMV & \Coloneqq & \cdots \mid \ECApNonMatchedAltAlt{\ECMV}
\end{array}\]
%
\begin{mathpar}
  \judgment{
    \ctxSynFixedInto{\ctx}{\EMV_1}{\ECMV_1}{\TMV} \\
    \notMatchedArrow{\TMV} \\\\
    \ctxAnaFixedInto{\ctx}{\EMV_2}{\ECMV_2}{\TUnknown}
  }{
    \ctxSynFixedInto{\ctx}{\EAp{\EMV_1}{\EMV_2}}{\ECAp{\ECApNonMatchedAltAlt{\ECMV_1}}{\ECMV_2}}{\TUnknown}
  }{ISAp2'}

  \judgment{
    \ctxSynType{\ctx}{\ECMV_1}{\TMV} \\
    \notMatchedArrow{\TMV} \\\\
    \ctxAnaType{\ctx}{\ECMV_2}{\TUnknown}
  }{
    \ctxSynType{\ctx}{\ECApNonMatchedAltAlt{\ECMV_1}{\ECMV_2}}{\TUnknown}
  }{MSAp2'}
\end{mathpar}

It is natural to extend this approach to other elimination forms that require the handling of
unmatched types, such as products. Another approach is to mark an error around the entire
application expression.

\subsubsection{Booleans}
\label{sec:calculus-booleans}

Boolean values are similarly straightforward:
%
\begin{mathpar}
  \judgment{ }{
    \ctxSynType{\ctx}{\ETrue}{\TBool}
  }{USTrue}

  \judgment{
  }{
    \ctxSynFixedInto{\ctx}{\ETrue}{\ECTrue}{\TBool}
  }{ISTrue}

  \judgment{ }{
    \ctxSynType{\ctx}{\ECTrue}{\TBool}
  }{MSTrue} \\

  \judgment{ }{
    \ctxSynType{\ctx}{\EFalse}{\TBool}
  }{USFalse}

  \judgment{
  }{
    \ctxSynFixedInto{\ctx}{\EFalse}{\ECFalse}{\TBool}
  }{ISFalse}

  \judgment{ }{
    \ctxSynType{\ctx}{\ECFalse}{\TBool}
  }{MSFalse}
\end{mathpar}

If-else expressions, however, offer another interesting case. In the unmarked language, we have
both explicit synthetic and analytic rules:
%
\begin{mathpar}
  \judgment{
    \ctxAnaType{\ctx}{\EMV_1}{\TBool} \\\\
    \ctxSynType{\ctx}{\EMV_2}{\TMV_1} \\
    \ctxSynType{\ctx}{\EMV_3}{\TMV_2}
  }{
    \ctxSynType{\ctx}{\EIf{\EMV_1}{\EMV_2}{\EMV_3}}{\TJoin{\TMV_1}{\TMV_2}}
  }{USIf}

  \judgment{
    \ctxAnaType{\ctx}{\EMV_1}{\TBool} \\\\
    \ctxAnaType{\ctx}{\EMV_1}{\TMV} \\
    \ctxAnaType{\ctx}{\EMV_2}{\TMV}
  }{
    \ctxAnaType{\ctx}{\ECIf{\EMV_1}{\EMV_2}{\EMV_3}}{\TMV}
  }{UAIf}
\end{mathpar}
%
In synthetic position, conditionals synthesize the \emph{join} of the branch types $\TMV_1$ and
$\TMV_2$, which we define inductively in \cref{fig:calculus-type-join}. In analytic position, since
there is an expected type for both branches, we shift the blame for any errors to them.

% TODO intuition why we use join

\newcommand{\joinsTo}[3]{\ensuremath{\TJoin{#1}{#2} & = & #3}}
\begin{figure}[htbp]
  \raggedright
  \judgbox{\ensuremath{\TJoin{\TMV_1}{\TMV_2}}} is a \emph{partial} metafunction defined as follows:
  %
  \[\begin{array}{rcl}
    \joinsTo{\TUnknown}{\TMV}{\TMV} \\
    \joinsTo{\TMV}{\TUnknown}{\TMV} \\
    \joinsTo{\TNum}{\TNum}{\TNum} \\
    \joinsTo{\TBool}{\TBool}{\TBool} \\
    \joinsTo{(\TArrow{\TMV_1}{\TMV_2})}{(\TArrow{\TMV_1'}{\TMV_2'})}{\TArrow{(\TJoin{\TMV_1}{\TMV_1'})}{(\TJoin{\TMV_2}{\TMV_2'})}} \\
  \end{array}\]
  %
  \caption{Type join.}
  \label{fig:calculus-type-join}
\end{figure}
% TODO intuition for type join?

Following \textsc{USIf} and \textsc{UAIf}, we may derive the following marking and typing rules.
Similar to previous cases, we simply recursively mark the guard and branch expressions and construct
the marked conditional from the results.
%
\begin{mathpar}
  \judgment{
    \ctxAnaFixedInto{\ctx}{\EMV_1}{\ECMV_1}{\TBool} \\\\
    \ctxSynFixedInto{\ctx}{\EMV_2}{\ECMV_2}{\TMV_1} \\
    \ctxSynFixedInto{\ctx}{\EMV_3}{\ECMV_3}{\TMV_2}
  }{
    \ctxSynFixedInto{\ctx}{\EIf{\EMV_1}{\EMV_2}{\EMV_3}}{\ECIf{\ECMV_1}{\ECMV_2}{\ECMV_3}}{\TJoin{\TMV_1}{\TMV_2}}
  }{ISIf}

  \judgment{
    \ctxAnaType{\ctx}{\ECMV_1}{\TBool} \\\\
    \ctxSynType{\ctx}{\ECMV_2}{\TMV_1} \\
    \ctxSynType{\ctx}{\ECMV_3}{\TMV_2}
  }{
    \ctxSynType{\ctx}{\ECIf{\ECMV_1}{\ECMV_2}{\ECMV_3}}{\TJoin{\TMV_1}{\TMV_2}}
  }{MSIf} \\

  \judgment{
    \ctxAnaFixedInto{\ctx}{\EMV_1}{\ECMV_1}{\TBool} \\\\
    \ctxAnaFixedInto{\ctx}{\EMV_2}{\ECMV_2}{\TMV} \\
    \ctxAnaFixedInto{\ctx}{\EMV_3}{\ECMV_3}{\TMV} \\
  }{
    \ctxAnaFixedInto{\ctx}{\EIf{\EMV_1}{\EMV_2}{\EMV_3}}{\ECIf{\ECMV_1}{\ECMV_2}{\ECMV_3}}{\TMV}
  }{IAIf}

  \judgment{
    \ctxAnaType{\ctx}{\ECMV_1}{\TBool} \\\\
    \ctxAnaType{\ctx}{\ECMV_1}{\TMV} \\
    \ctxAnaType{\ctx}{\ECMV_2}{\TMV}
  }{
    \ctxAnaType{\ctx}{\ECIf{\ECMV_1}{\ECMV_2}{\ECMV_3}}{\TMV}
  }{MAIf}
\end{mathpar}

However, in synthetic mode, it may be the case that the two branch types have no join. This occurs,
in fact, when they are inconsistent. This motivates \emph{the inconsistent branches mark}
$\ECInconBr{\ECMV}{\ECMV}{\ECMV}$. Then, adding the following rules ensures marking totality on
conditionals:
%
\begin{mathpar}
  \judgment{
    \ctxAnaFixedInto{\ctx}{\EMV_1}{\ECMV_1}{\TBool} \\
    \ctxSynFixedInto{\ctx}{\EMV_2}{\ECMV_2}{\TMV_1} \\\\
    \ctxSynFixedInto{\ctx}{\EMV_3}{\ECMV_3}{\TMV_2} \\
    \inconsistent{\TMV_1}{\TMV_2}
  }{
    \ctxSynFixedInto{\ctx}{\EIf{\EMV_1}{\EMV_2}{\EMV_3}}{\ECInconBr{\ECMV_1}{\ECMV_2}{\ECMV_3}}{\TUnknown}
  }{ISInconsistentBranches}

  \judgment{
    \ctxAnaType{\ctx}{\ECMV_1}{\TBool} \\
    \ctxSynType{\ctx}{\ECMV_2}{\TMV_1} \\\\
    \ctxSynType{\ctx}{\ECMV_3}{\TMV_2} \\
    \inconsistent{\TMV_1}{\TMV_2}
  }{
    \ctxSynType{\ctx}{\ECInconBr{\ECMV_1}{\ECMV_2}{\ECMV_3}}{\TUnknown}
  }{MSInconsistentBranches}
\end{mathpar}

\subsubsection{Subsumption}
\label{sec:calculus-subsumption}

>>>>

\subsection{Let with Composite Patterns}
\label{sec:calculus-let}

\begin{figure}[htbp]
  \raggedright
  \judgbox{\ensuremath{\ctxSynType{\Gamma_{in}}{p}{\tau ; \Gamma_{out}}}} $\TMV$ has matched binary product type $\TPair{\TMV_1}{\TMV_2}$
  %
  \begin{mathpar}
    \judgment{ }{
      \matchedPair{\TUnknown}{\TUnknown}{\TUnknown}
    }{TMPHole}
